Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-11 conference-materials/2025_posit-conf/draft.qmd
Enhance revealjs presentation with behavioral design refinements
- Added `revealjs` format to `_quarto.yml` with shared custom CSS for unified styling
- Updated slide HTML and QMD sources for layout improvements, dark theme backgrounds, and title block visuals
- Rewrote notes for clarity and alignment with storytelling flow
- Improved user experience through progressive disclosure, image layouts, and text hierarchy
- Updated `renv.lock` to reflect `bidux` source now hosted on CRAN instead of GitHub

---
title: "Death by Dropdown? Engineer Insightful Shiny Apps with Behavioral Science"
subtitle: "posit::conf(2025)"
author: "Jeremy Winget, PhD"
date: "2025-09-17"
execute:
  echo: true
  eval: true
format:
  revealjs:
    slide-number: true
    transition: fade
    footer: "posit::conf(2025) | jrwinget\\.com"
    preview-links: auto
    code-block-border-left: true
    highlight-style: github
    auto-animate: true
    chalkboard: false
    width: 1600
    height: 900
---

## {.center background-color="#2c3e50" auto-animate=true}

::: {.r-fit-text}
**Ever open a dashboard with 18 filters across 6 tabs...<br>
and immediately get the urge to close your laptop?**
:::

:::: {.fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="60%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
::::

::: {.notes}
- 10 seconds
- Ever open a dashboard with 18 filters across 6 tabs...and just get the urge to immediately close your laptop?
- [CLICK] ....or worse
- [PAUSE] I get the feeling I'm not alone...(OR maybe I am)
- "Welcome to dashboard therapy"
:::

@@

:::: {.columns}
::: {.column width="50%" .text-xl}
### Quick show of hands:
- Built Shiny app 10+ filters?
- User asked "one more dropdown"?
- More `selectInput()` than insights?
:::

::: {.column width="50%" .fragment .text-xl}
### Welcome! You're among friends

👋 **Hi, I'm Jeremy! Social psychologist turned full stack engineer**<br>
*Think of our time together as a little therapy session for dropdown overload*
:::
::::

@@

::: {.notes}
- 20 seconds
- Raise hand if you've built a Shiny app with 10+ filters
- Or if you've had a user ask for "just one more dropdown"
- Or if you've added more `selectInput()`s to your dashboard than actual insights
- [CLICK] "Well if so...let me welcome you to my dashboard support group"
- You're among friends here
- [PAUSE] Hi, I'm Jeremy! ...social psych turned engineer
- Think of our time together today as a little therapy session for dropdown overload
:::

## Why Listen to Me? {background-color="#2c3e50"}

:::: {.columns}
::: {.column width="50%"}
**PhD in Applied Psychology**<br>
*How people actually process information*

**Created {bidux} package**<br>
*Your behavioral scientist in the console*
:::

::: {.column width="50%" .fragment}
### Core Realization

> "Most dashboard failures aren't technical. They're psychological."
:::
::::

::: {.notes}
- 30 seconds
- "PhD in Applied Psych...studied how people/groups process information and make decisions"
- Transitioned from academia to industry...where I got involved leading Shiny teams at scale
- "Have nearly 10 years experience building enterprise apps/dashboards folks actually use"
- But it wasn't always that way...I've built MANY dashboards that flopped even though they were technically sound
- This is what led to my core realization: Most dashboard failures aren't technical, they're psychological
- "Let me show you what I mean"
:::

@@

::: {.notes}
- 45 seconds
- "I worked with a client who had this 'comprehensive' dashboard"
- Blinded here for confidentiality...but incredibly close to the actual design
- "18 filters, 6 tabs, every possible view, nearly zero adoption"
- The client was frustrated bc they had spent months building this thing, but no one used it
- I had some immediate thoughts, but being a behavioral scientist, I wanted to validate this
    1. So that the client's time and efforts weren't wasted
    2. And so I could show them data to back up my claims and explain why these weren't just subjective issues
- [CLICK] "So we ran telemetry for 30 days..."
:::

@@

## Enter: Behavioral Science {background-color="#3498db"}

@@

## The Solution: Your Dashboard Journey 🚗 {background-color="#34495e"}

@@

:::: {.columns}
::: {.column width="40%"}
### Think of it like a road trip

📍 **Interpret** where you're going<br>
⚠️ **Notice** the warning lights<br>
🌦️ **Anticipate** weather changes<br>
🗺️ **Structure** the best route<br>
✅ **Validate** safe arrival
:::

::: {.column width="40%" .fragment}
### The BID Framework

📍 **Interpret** user needs<br>
⚠️ **Notice** friction points<br>
🌦️ **Anticipate** cognitive biases<br>
🗺️ **Structure** information flow<br>
✅ **Validate** understanding
:::

::: {.column width="20%"}
![](img/bid-framework.png){width="100%"}
:::
::::

@@

::: {.notes}
- 45 seconds
- "Think of building dashboards like you would a road trip"
- "You need to interpret where you're going, notice warning lights..."
- [CLICK] "This is the BID framework"
- "Five *iterative* stages to prevent dashboard disasters"
:::

@@

## INTERPRET: Start with Why {background-color="#3498db"}

@@

::: {.notes}
- 1 minute
- "Start with why - what question does this dashboard answer?"
- [CLICK] "10 word rule"
- Why this is important - most devs think they do this, but telemetry says otherwise. Most jump straight to filters.
- But as devs and data scientists, we're not the best at this. We think we know what users want...and if we're not sure, we just add more filters/options
- But this is exactly what leads to cognitive overload
- `bid_interpret` helps you clarify purpose, user needs, and data story...and if you don't know any of those elements, it prompts you to think about them (which might be a good time to pause and actually chat with your users or client)
- On the back end, it's creating an object that can be used later in the process to make smart suggestions for your specific users
:::

## NOTICE: Find the Core Problems {background-color="#e74c3c"}

@@

::: {.notes}
- 1 minute
- "Notice finds the concepts behind the friction points"
- "Minimal input gives maximum insight"
- Auto-suggests behavioral theories...or power users can override
- "Look at this heatmap - 14 filters completely ignored"
- "This is choice overload in action"
:::

## ANTICIPATE: Guard Against Biases {background-color="#f39c12"}

@@

## STRUCTURE: Progressive Disclosure {background-color="#16a085"}

@@

## VALIDATE: Plain Language Wins {background-color="#27ae60"}

@@

## The Complete Workflow {background-color="#8e44ad"}

@@

## Real Impact Across Industries {background-color="#27ae60"}

@@

::: {.notes}
- 30 seconds
- "Real results from 12 enterprise deployments"
- "82% reduction in time to insight"
- "143% improvement in task completion"
- "This works across industries"
:::

## Your Turn: Start Today {.center background-color="#3498db"}

@@

@@

::: {.column width="30%" .fragment}
*Special thanks to the posit::conf organizers,<br>
the Shiny community,<br>
and all of you for caring about your users*
:::

::: {.column width="30%"}
![](img/hex-bidux.png){class="hex-logo" width="200"}
:::
::::

>>>>

---
title: "Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science"
author: "Jeremy Winget, PhD"
date: "September 17, 2025"
date-format: "MMMM D, YYYY"
execute:
  echo: true
  eval: true
format:
  revealjs:
    code-block-background: true
    highlight-style: github
    slide-number: false
    transition: fade
    auto-animate: true
    width: 1600
    height: 900
    footer: "posit::conf(2025) | jrwinget\\.com"
    title-slide-attributes:
      data-background-image: "img/posit-conf.png"
      data-background-position: "top left"
      data-background-repeat: "no-repeat"
      data-background-color: "#213d4f"
      data-background-opacity: "0.25"
      data-background-size: "45%"
---

## {.center background-color="#2c3e50"}

::: {.r-fit-text}
**Ever open a dashboard with 18 filters across 6 tabs...<br>
and wonder where to even start?**
:::

:::: {.columns .fragment}
::: {.column width="45%"}
<div class="tenor-gif-embed" data-postid="596341483704119392" data-share-method="host" data-aspect-ratio="1.32447" data-width="50%"><a href="https://tenor.com/view/monkey-computer-work-bug-developer-gif-596341483704119392">Monkey Computer GIF</a>from <a href="https://tenor.com/search/monkey-gifs">Monkey GIFs</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::

::: {.column width="45%"}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="60%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

::: {.notes}
- 10 seconds
- Ever open a dashboard with 18 filters across 6 separate tabs...and try to figure out where to even start?
- I mean, that's a lot of data and different ways slicing it to get a full understanding of what's going on...so what do you do?
- That is...if you haven't made your career out of exploring data sets like most executives making decisions...what do you do?
- [CLICK] Maybe you do something like this monkey....just start clicking randomly hoping something sticks
- [PAUSE]
- [CLICK] Or maybe your feelings are a bit more intense...
- I get the feeling I'm not the only one who's been here before
:::

@@

:::: {.columns}
::: {.column width="50%" .text-xl}
### Raise hand if you've ever:
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?

::: {.fragment}
### Welcome! You're among friends

*Think of our time together as a little therapy session for dropdown overload*
:::
:::

::: {.column width="50%" .fragment .text-xl}
### 👋 **Hi, I'm Jeremy!

**PhD in Applied Psychology**<br>
*How people actually process information*

**Created {bidux} package**<br>
*Your behavioral scientist in the console*

I'm a psychologist by training — but not the kind you're thinking of.**<br>

*Contrary to popular belief, most psychologists work with data, not clients.*<br>

**Academic → Tech trajectory**: Applied psychology PhD → Full stack engineer
:::
::::

::: {.fragment}
### Core Realization

> "Most dashboard failures aren't technical. They're psychological."
:::

@@

::: {.notes}
- 1 minute
- Quick show of hands...who's created a user interface with more than 10 filters?
- Or if ever had a client/user ask for "just one more dropdown"
- Or built a Shiny app with more `selectInput()`s than insights
- [CLICK] If so...welcome you to my dashboard support group
- Think of our time together today as a little therapy session for dropdown overload
- [PAUSE] Hi, I'm Jeremy! I'm a psychologist by training...but not the kind you're thinking of!
- Contrary to popular belief...there are many types of psychologists...and most of us with data, not clients!
- Clarify the therapy metaphor: support group hook connects our shared pain to therapeutic framing
- Most psychologists actually work with data, studying behavior patterns, not doing individual therapy
- My path: academic research → tech industry, bringing behavioral insights to engineering challenges
- "PhD in Applied Psych...studied how people/groups process information and make decisions"
- Transitioned from academia to industry...where I got involved leading Shiny teams at scale
- "Have nearly 10 years experience building enterprise apps/dashboards folks actually use"
- But it wasn't always that way...I've built MANY dashboards that flopped even though they were technically sound
- This is what led to my core realization: Most dashboard failures aren't technical, they're psychological
- "Let me show you what I mean"
:::

@@

::: {.notes}
- 45 seconds
- "I worked with a client who had this 'comprehensive' dashboard"
- Blinded here for confidentiality...but incredibly close to the actual design
- "18 filters, 6 tabs, every possible view, nearly zero adoption"
- Why this matters: This represents thousands of developer hours and client budget wasted on technically perfect but psychologically broken software
- The client was frustrated bc they had spent months building this thing, but no one used it
- As developers, we can relate - we've all built features nobody uses. But here's why this matters more than frustration:
    1. Technical debt from unused complexity compounds over time
    2. User abandonment patterns replicate across similar projects
    3. Without data-driven insights, we keep making the same UX mistakes
- [CLICK] "So we ran telemetry for 30 days..." - This data changed everything about how I approach dashboard design
:::

@@

## Enter: Behavioral Science {background-color="#5b6bbf"}

@@

## Classic Example: The Paradox of Choice {background-color="#5b6bbf"}

:::: {.columns}
::: {.column width="50%"}
![Jam Study Results](img/jam-study-behavioral.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::

::: {.column width="50%"}
### Sheena Iyengar's Jam Study

::: {.text-large}
**24 varieties displayed**: 60% stopped to look<br>
**6 varieties displayed**: 40% stopped to look

But...

**24 varieties**: Only 3% bought jam<br>
**6 varieties**: 30% bought jam
:::

::: {.fragment .text-xl style="color: var(--coral); font-weight: 600; margin-top: 30px;"}
**10x higher conversion with fewer options**
:::
:::
::::

::: {.notes}
- 45 seconds
- "Classic behavioral science example that perfectly mirrors our dashboard problem"
- "Sheena Iyengar's famous jam study at Columbia University"
- Show image: Side-by-side grocery store displays - one with 24 jam varieties (crowded, overwhelming), one with 6 varieties (clean, organized). Include simple bar charts showing the dramatic difference in purchase rates.
- "More choice attracted browsers, but paralyzed buyers"
- [CLICK] "This is exactly what happens with dashboard filters - more options kill engagement"
- "This principle drives everything we'll build with the BID framework"
:::

## The Solution: Your Dashboard Journey 🚗 {background-color="#2c3e50"}

@@

:::: {.columns}
::: {.column width="30%"}
![Road Trip Adventure](img/roadtrip-journey.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::

::: {.column width="35%"}
### Think of it like a road trip

📍 **Interpret** where you're going<br>
⚠️ **Notice** the warning lights<br>
🌦️ **Anticipate** weather changes<br>
🗺️ **Structure** the best route<br>
✅ **Validate** safe arrival

::: {.fragment}
![VW Bus Spirit](img/vw-bus-adventure.png){width="80%" style="margin-top: 20px; border-radius: 8px;"}
:::
:::

::: {.column width="35%" .fragment}
### The BID Framework

📍 **Interpret** user needs<br>
⚠️ **Notice** friction points<br>
🌦️ **Anticipate** cognitive biases<br>
🗺️ **Structure** information flow<br>
✅ **Validate** understanding

![BID Hex](img/bid-framework.png){width="80%" style="margin-top: 20px;"}
:::
::::

@@

::: {.notes}
- 45 seconds
- Show image: Scenic road stretching toward mountains with modern car, representing the journey toward better UX. Should evoke adventure, possibility, and structured progress.
- "Think of building dashboards like you would a road trip"
- "You need to interpret where you're going, notice warning lights..."
- [CLICK] Show image: Classic VW bus with hex-style stickers, embodying the "long strange trip" philosophy of exploration and community. Connects to Grateful Dead culture of innovation and shared journey.
- [CLICK] "This is the BID framework"
- "Five *iterative* stages to prevent dashboard disasters"
- "Like any good road trip, it's not just about the destination - it's about having the right framework to explore and discover along the way"
:::

@@

## INTERPRET: Start with Why {background-color="#5b6bbf"}

@@

::: {.notes}
- 1 minute
- "Start with why - what question does this dashboard answer?"
- [CLICK] "10 word rule"
- Why this matters for your daily work: Clear purpose drives every technical decision you make
- Most devs think they do this, but telemetry says otherwise. We jump straight to technical implementation
- The cost: Every filter you add without purpose increases cognitive load exponentially
- This isn't just UX theory - it's engineering efficiency. When you clarify purpose first:
    1. You write less code because you know what to exclude
    2. Your data queries are more focused and performant
    3. User testing becomes objective rather than subjective
    4. You can defend design decisions with evidence, not opinions
- `bid_interpret` creates a structured object that feeds into later stages, making behavioral science actionable in your R workflow
- Think of it as writing better requirements that actually connect to user psychology
:::

## NOTICE: Find the Core Problems {background-color="#5b6bbf"}

@@

::: {.notes}
- 1 minute
- "Notice finds the concepts behind the friction points"
- Why this stage matters for developers: It turns user complaints into actionable engineering insights
- Look at this heatmap - 14 filters completely ignored. This isn't just usage data, it's technical debt visualization
- What this means for your codebase:
    1. Those unused filters are maintenance overhead with zero ROI
    2. Database queries are running for features nobody touches
    3. UI complexity is slowing render times without value
    4. Future feature requests will compound these problems
- `bid_notice` auto-suggests behavioral theories to explain patterns, or power users can override
- This transforms "users don't like it" into "choice overload is killing task completion" - now you have a specific problem to solve with code
:::

## ANTICIPATE: Guard Against Biases {background-color="#5b6bbf"}

@@

## STRUCTURE: Progressive Disclosure {background-color="#5b6bbf"}

@@

## VALIDATE: Plain Language Wins {background-color="#5b6bbf"}

@@

## The Complete Workflow {background-color="#2c3e50"}

@@

## Real Impact Across Industries {background-color="#5b6bbf"}

@@

::: {.notes}
- 30 seconds
- "Real results from 12 enterprise deployments"
- Why these metrics matter for your engineering career:
    1. 82% reduction in time to insight = users find value faster = better adoption metrics for your projects
    2. 143% improvement in task completion = fewer support tickets = less maintenance overhead
    3. These improvements compound: happy users request thoughtful features, not band-aid fixes
- "This works across industries" - the psychological principles are universal, but implementation adapts to your specific R/Shiny tech stack
- Bottom line: Better user psychology = better engineering outcomes = more impactful projects in your portfolio
:::

## Your Turn: Start Today {.center background-color="#5b6bbf"}

@@

::: {.column width="30%"}
![Community Spirit](img/posit-conf-community.png){width="100%" style="border-radius: 8px; opacity: 0.8;"}
:::

@@

::: {.column width="30%"}
![BID Hex](img/hex-bidux.png){class="hex-logo" width="200"}

::: {.fragment .text-small style="margin-top: 20px; color: var(--warm-white);"}
*Special thanks to the posit::conf organizers,<br>
the Shiny community,<br>
and all of you for caring about your users*
:::
:::
::::
